---
title: "Getting Started with myceliumr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with myceliumr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

myceliumr provides a lightweight framework for managing cross-species study data (rat, mouse, human) with validation, ortholog support, and standardized storage for WES and snRNA-seq outputs.

## Installation

```r
devtools::install_github("samuelbharti/myceliumr")
```

## Quick Start

Load the package and example data:

```{r}
library(myceliumr)
data(example_cohort)
```

## Create a Study

A Study object describes the overall research project:

```{r}
study <- study_new(
  study_id = "STUDY001",
  title = "My Genomics Project",
  description = "Comparing rat and mouse expression",
  hypotheses = "Orthologs show conserved patterns",
  aims = "Map regulatory regions",
  assays = c("WES", "snRNA-seq"),
  genome_builds = list(rat = "rn6", mouse = "mm10")
)

study
```

## Create Subjects

Individual subjects (animals or samples):

```{r}
subject1 <- subject_new(
  subject_id = "RAT001",
  species = "rat",
  sex = "M",
  strain = "Lewis",
  genotype = "WT"
)

subject1
```

Supported species: `rat`, `mouse`, `human`.

## Create a Manifest

A manifest CSV defines all subjects and sample mappings. Required columns:

- `subject_id` - unique identifier
- `species` - rat, mouse, or human

Optional columns: `sex`, `strain`, `genotype`, `cohort`, `timepoint`, `notes`, and assay IDs like `assay_wes_id`, `assay_snrna_id`.

Example:

```{r}
manifest <- tibble::tribble(
  ~subject_id, ~species, ~sex, ~strain, ~genotype, ~cohort, ~assay_wes_id,
  "RAT001", "rat", "M", "Lewis", "WT", "Control", "WES_R001",
  "RAT002", "rat", "F", "Lewis", "WT", "Control", "WES_R002",
  "MOUSE001", "mouse", "M", "C57BL/6", "WT", "Control", "WES_M001",
  "MOUSE002", "mouse", "F", "C57BL/6", "KO", "Treatment", "WES_M002"
)

manifest
```

Or load from CSV:

```r
manifest <- read_manifest_csv("path/to/manifest.csv")
```

## Create a Cohort

A Cohort combines a Study with validated subject data:

```{r}
# First validate the manifest
manifest_split <- validate_manifest(manifest)

# Create cohort from validated components
cohort <- cohort_new(
  study = study,
  subject_tbl = manifest_split$subject_tbl,
  sample_map = manifest_split$sample_map
)

cohort
```

## Access Cohort Data

Get the subject table:

```{r}
subjects <- cohort@subject_tbl
subjects
```

Get the sample mapping:

```{r}
samples <- cohort@sample_map
samples
```

## Validation

Validation happens automatically during object creation. Invalid data raises clear error messages:

```r
# This will fail - invalid species
bad_subject <- subject_new(
  subject_id = "BAD001",
  species = "zebra fish"  # Error: not rat/mouse/human
)

# This will fail - missing required columns
bad_manifest <- tibble::tibble(subject_id = c("S1", "S2"))
cohort_bad <- cohort_new(
  study = study,
  subject_tbl = bad_manifest,
  sample_map = tibble::tibble()
)
# Error: species column is required
```

## Example Dataset

Use the built-in example cohort to explore:

```{r}
data(example_cohort)

# View the study
example_cohort@study

# View subjects
example_cohort@subject_tbl

# View samples
example_cohort@sample_map
```

## Next Steps

- Define your study metadata with `study_new()`
- Prepare a manifest CSV with subject and sample info
- Create a Cohort to validate and manage your data
- Use `@` to access cohort components for downstream analysis
